---
title: "R Notebook"
output: html_notebook
---

10강에서 사용한 유방암 환자 데이터를 불러온다. 

```{r}
library(tidyverse)
brcasample<-read_tsv('BRCA_clinicalMatrix.dms')

#암 환자 RNA-Seq 데이터를 불러온다. 
brca_cancer <-read_tsv("brca-rsem-count-tcga-t.txt")

cancer <- colnames(brca_cancer)[3:length(colnames(brca_cancer))]
samplelist <- tibble (
             sample = cancer,
             condition="T"
)

```
환자 데이터를 불러오고, 필요한 정보만을 추린다.  환자 샘플 데이터, 무진행 생존기간, HER2 status, PR status, ER status, Metastasis status

```{r}
samplelist$patient = substr(samplelist$sample,1,12)
cancersample <- brcasample %>% inner_join(samplelist, by=c("_PATIENT"="patient")) %>%
                select('sample','OS_Time_nature2012','HER2_Final_Status_nature2012','PR_Status_nature2012','ER_Status_nature2012','Metastasis_nature2012') %>%
                drop_na('OS_Time_nature2012','HER2_Final_Status_nature2012','PR_Status_nature2012','ER_Status_nature2012','Metastasis_nature2012')
  
cancersample
```

샘플 데이터가 있는 데이터만 추린다. 

```{r}
brca <- brca_cancer %>% select (cancersample$sample) 
rownames(brca) <- brca_cancer$Hugo_Symbol
```
CPM 을 계산하고, CPM이 10 이 넘는 환자 샘플이 10명 이상인 유전자만 남긴다. 
이를  dat 에 저장한다. 

```{r}
library(edgeR)
library(WGCNA)

cpmbrca <- cpm(brca)

keep <- rowSums(cpmbrca>10) >= 10
cp <- cpmbrca[keep,]
dat <- t(cp)

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

유전자의 발현을 이용하여 Tree를 그려 보자. 

```{r}
sampleTree = hclust(dist(dat), method = "average");
```

```{r, fig.width = 20, fig.height = 12}
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, 
     cex.axis = 1.5, cex.main = 2)

# Plot a line to show the cut
abline(h = 70000, col = "red");
```

지나친 outlier인 샘플을 1개 삭제한다. 1개의 샘플이 삭제되고, 699명의 샘플이 남는다. 

```{r}
clust = cutreeStatic(sampleTree, cutHeight = 70000, minSize = 10)
table(clust)
# clust 1 contains the samples we want to keep.
keepSamples = (clust==1)
datExpr = dat[keepSamples, ]
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)
```
샘플 
```{r}
rownames(datExpr)
```


환자 샘플을 정리하여 다음과 같은 포맷으로 바꾼다. 

```{r}
cs <- cancersample %>% filter (sample %in% rownames(datExpr)) %>%
                 column_to_rownames(var="sample")
cs$HER2_Final_Status_nature2012<-as.numeric(as.factor(cs$HER2_Final_Status_nature2012))
cs$PR_Status_nature2012<-as.numeric(as.factor(cs$PR_Status_nature2012))
cs$ER_Status_nature2012<-as.numeric(as.factor(cs$ER_Status_nature2012))
cs$Metastasis_nature2012<-as.numeric(as.factor(cs$Metastasis_nature2012))
cs
```
환자 샘플의 트리와 표현형을 그린다. 

```{r, fig.width = 30, fig.height = 12}

pdf('Tree.pdf',width=13,height=13)
par(mai=c(6,6,6,6))
# Re-cluster samples
sampleTree2 = hclust(dist(datExpr), method = "average")
# Convert traits to a color representation: white means low, red means high, grey means missing entry
traitColors = numbers2colors(cs, signed = FALSE);
# Plot the sample dendrogram and the colors underneath.
plotDendroAndColors(sampleTree2, traitColors,
                    groupLabels = names(cs), 
                    main = "Sample dendrogram and trait heatmap")
dev.off()

```
네트워크 
```{r}
options(stringsAsFactors = FALSE);
# Allow multi-threading within WGCNA. At present this call is necessary.
# Any error here may be ignored but you may want to update WGCNA if you see one.
# Caution: skip this line if you run RStudio or other third-party R environments.
# See note above.
allowWGCNAThreads()
# Load the data saved in the first part

#=====================================================================================
#
#  Code chunk 2
#
#=====================================================================================


# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
# Plot the results:
sizeGrWindow(15, 10)
par(mfrow = c(1,2));
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")

```

네트워크의 급수를 4로 설정 

```{r}

softPower = 4;
adjacency = adjacency(datExpr, power = softPower);

# Turn adjacency into topological overlap
TOM = TOMsimilarity(adjacency);
dissTOM = 1-TOM


# Call the hierarchical clustering function
geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
sizeGrWindow(12,9)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04);

```

모듈을 색으로 나눔 

```{r}


# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30;
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
                            deepSplit = 2, pamRespectsDendro = FALSE,
                            minClusterSize = minModuleSize);
table(dynamicMods)


# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
sizeGrWindow(8,6)
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")


```

모듈을 통합하여 소수의 모듈로 줄이자. 

```{r, fig.width = 12, fig.height = 12}


# Calculate eigengenes
MEList = moduleEigengenes(datExpr, colors = dynamicColors)
MEs = MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
# Plot the result
sizeGrWindow(7, 6)
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")

MEDissThres = 0.4
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")
# Call an automatic merging function
merge = mergeCloseModules(datExpr, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors;
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs;
```

```{r}

sizeGrWindow(12, 9)
#pdf(file = "Plots/geneDendro-3.pdf", wi = 9, he = 6)
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
#dev.off()

# Rename to moduleColors
moduleColors = mergedColors
# Construct numerical labels corresponding to the colors
colorOrder = c("grey", standardColors(50));
moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs;
# Save module colors and labels for use in subsequent parts
save(MEs, moduleLabels, moduleColors, geneTree, file = "FemaleLiver-02-networkConstruction-stepByStep.RData")

```
모듈과 표현형과의 관계 
```{r, fig.width = 12, fig.height = 12}


# Define numbers of genes and samples
nGenes = ncol(datExpr);
nSamples = nrow(datExpr);
# Recalculate MEs with color labels
MEs0 = moduleEigengenes(datExpr, moduleColors)$eigengenes
MEs = orderMEs(MEs0)
moduleTraitCor = cor(MEs, cs, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);

sizeGrWindow(20,20)
# Will display correlations and their p-values
textMatrix =  paste(signif(moduleTraitCor, 2), "\n(",
                    signif(moduleTraitPvalue, 1), ")", sep = "");
dim(textMatrix) = dim(moduleTraitCor)

# Display the correlation values within a heatmap plot
pdf('Module-Trait.pdf',width=13,height=13)
par(mar = c(9, 9, 9, 9));
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(cs),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = greenWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships"))

dev.off()
```

HER2 phenotype과 모듈과의 관계 

```{r}
# Define variable weight containing the weight column of datTrait
her2 = as.data.frame(cs$HER2_Final_Status_nature2012);
names(her2) = "her2"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificance = as.data.frame(cor(datExpr, her2, use = "p"));
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance), nSamples));

names(geneTraitSignificance) = paste("GS.", names(her2), sep="");
names(GSPvalue) = paste("p.GS.", names(her2), sep="");


module = "salmon4"
column = match(module, modNames);
moduleGenes = moduleColors==module;

sizeGrWindow(7, 7);
par(mfrow = c(1,1));
verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                   abs(geneTraitSignificance[moduleGenes, 1]),
                   xlab = paste("Module Membership in", module, "module"),
                   ylab = "Gene significance for her2",
                   main = paste("Module membership vs. gene significance\n"),
                   cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)

```
Salmon4 모듈에 있는 유전자는? 
```{r}
salmon4 <- geneModuleMembership[moduleGenes,]
salmon4 <- salmon4 %>% rownames_to_column(var='gene') %>%
            dplyr::select (gene,MMsalmon4)
gs <- geneTraitSignificance %>% rownames_to_column(var='gene') %>%
      inner_join(salmon4, by='gene')
gs
```

```{r}
TOM = TOMsimilarityFromExpr(datExpr, power = 4);
```

```{r}
install.packages('igraph')
```


```{r, fig.width = 10, fig.height = 10}
library(igraph)
modules <- c("salmon4")
probes <- rownames(geneModuleMembership)
inModule <- (moduleColors==modules)
modProbes <- probes[inModule]
modTOM <- TOM[inModule,inModule]
dimnames(modTOM) <- list(modProbes,modProbes)
graph<-graph.adjacency(modTOM,mode='directed',weighted=TRUE)
graph<-simplify(graph)
par(mar=c(0,0,0,0))
#graph<- delete.vertices(graph, degree(graph)==0)
plot(graph,
     layout=layout.fruchterman.reingold(graph, weights=E(graph)$weight),
     edge.arrow.size=0.1
     )
```


ER Response관련 모듈 

```{r}
er2 = as.data.frame(cs$ER_Status_nature2012);
names(er2) = "er2"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificance = as.data.frame(cor(datExpr, er2, use = "p"));
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance), nSamples));

names(geneTraitSignificance) = paste("GS.", names(er2), sep="");
names(GSPvalue) = paste("p.GS.", names(er2), sep="");

module = "green"
column = match(module, modNames);
moduleGenes = moduleColors==module;

sizeGrWindow(7, 7);
par(mfrow = c(1,1));
verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                   abs(geneTraitSignificance[moduleGenes, 1]),
                   xlab = paste("Module Membership in", module, "module"),
                   ylab = "Gene significance for ER",
                   main = paste("Module membership vs. gene significance\n"),
                   cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)

```

```{r}
cyan <- geneModuleMembership[moduleGenes,]
cyan <- cyan %>% rownames_to_column(var='gene') %>%
            dplyr::select (gene,MMcyan)
gs <- geneTraitSignificance %>% rownames_to_column(var='gene') %>%
      inner_join(cyan, by='gene')
```

Cyan 모듈의 유전자들의 Go term 분석 

```{r}
library(goseq)
genes <- as.integer(moduleGenes)
names(genes)<-rownames(geneModuleMembership)
pwf <- nullp(genes,"hg19","geneSymbol")
GO.wall <- goseq(pwf,"hg19","geneSymbol")
GO.wall$FDR <-p.adjust(GO.wall$over_represented_pvalue, method="fdr")

```

